Limited physiological plasticity and reduced relative variance in physiological rates of ectotherm populations under climate change

### 2024-07-04 ms.qmd
UPdated and working on LH secin in discussion

We show that acclimation of physiological rates and changes in relative variance differ across habitats. Our findings confirm previous results that quantify the different capacity of terrestrial, marine and freshwater ectotherms to acclimate [@Seebacher2015-wi; @Gunderson2015-hz; @morley2019physiological]. Our analysis generally supports that of @Seebacher2015-wi, @Gunderson2015-hz and @morley2019physiological that all show a general inability of terrestrial ectotherms to physiologically acclimate. The change in acclimation $Q_{10}$ we found was similar to @Seebacher2015-wi for freshwater organisms (~17%), but higher in marine ectotherms (decrease of 16% versus ~10% in @Seebacher2015-wi), and opposite in terrestrial ectotherms (increase of ~6% compared to an ~8% decrease in @Seebacher2015-wi), possibly due to the use of new $Q_{10}$ based effect sizes that allow us to control for sampling variance. Importantly, while marine and freshwater ectotherms were capable of partial acclimation, the effect size was small (amounting to $Q_{10}$ dropping from ~1.8 to 1.6).

@@

Our results further highlight the potential vulnerability of terrestrial ectotherms to climate change, while calling into question if acclimation in marine systems will be able to counteract the potentially negative effects of reductions in relative variance. Assuming changes in relative variation in physiological rates are underpinned by genetic variation, and there is a genetic correlation with fitness, such reductions in physiological variance could limit a populations ability to adapt to future climates [@urban2023and]. Of course, considering how behavioural modifications buffer organisms from high temperatures is also needed. For example, terrestrial species have greater opportunity for behavioural microhabitat selection which is expected to decrease the impacts of climate change in thermally heterogeneous environments [@huey2012predicting]. Importantly, while we seen a reduction in relative variation, absolute variation in physiological rates did increase with temperature (@fig-s1) -- albeit to a much lesser extent in terrestrial and marine habitats compared to freshwater -- supporting the general notion that capacity for adaptation may still be limited in terrestrial and marine habitats. As of yet, it is unclear whether a focus on relative or absolute variance should be prioritized. Future theoretical models, that vary both mean and variance, are needed to characterise the interplay between changes in mean and variance.

### ***Plasticity and relative variance in physiological rates do not differ between life stages***
Life-history stages often occupy different ecological niches and exhibit different physiological responses, levels of plasticity, and patterns of mortality in response to temperature. As such, it is becoming increasingly important to understand how climate change will impact different life-history stages [@levy2015resolving; @petitgas2013impacts]. Greater plasticity in early life stages of development is expected to be important to increase resilience to the effects of climate change because early life stages are often particularly vulnerable periods in development [@martin2015age; @stearns1976life]. We show that there were generally similar patterns between early and late life stages across a diversity of taxa, both in terms of variance changes and the capacity for plasticity. On average, our results show that early life stages are not more vulnerable to the impacts of climate change. However, our general meta-analytic findings do not imply that early life stages are not always less plastic, but that such responses are likely context or trait dependent [@carter2020life; @moghadam2019heat]. For example, @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults. In contrast, @carter2020life showed greater thermal plasticity in metabolic rate for adults but little in pupae of *Onthophagus taurus*.  Despite variation in the literature, very few studies measure multiple life-history stages within the same population. In future research, there needs to be more studies that explicitly compare differences in plasticity among life stages across multiple traits within the same population.

>>>>

We show that acclimation of physiological rates and changes in relative variance differ across habitats. Our findings confirm previous results that quantify the different capacity of terrestrial, marine and freshwater ectotherms to acclimate [@Seebacher2015-wi; @Gunderson2015-hz; @morley2019physiological]. Our analysis generally supports that of @Seebacher2015-wi, @Gunderson2015-hz and @morley2019physiological that all show a general inability of terrestrial ectotherms to physiologically acclimate. The change in acclimation $Q_{10}$ we found was similar to @Seebacher2015-wi for freshwater organisms (~17%), but higher in marine ectotherms (decrease of 16% versus ~10% in @Seebacher2015-wi), and opposite in terrestrial ectotherms (increase of ~6% compared to an ~8% decrease in @Seebacher2015-wi), possibly due to the use of new $Q_{10}$ based effect sizes that allow us to control for sampling variance. Greater capacity for acclimation in aquatic organisms may be the result of fewer opportunities for behaviuoral thermoregulation in aquatic environments making physiological remodeling important for maintaining homeostasis [@Gunderson2015-hz; @morley2019physiological]. While marine and freshwater ectotherms were capable of partial acclimation it is still important to recognise that, on average, the effect size was small (amounting to $Q_{10}$ dropping from ~1.8 to 1.6).

@@

Our results further highlight the potential vulnerability of terrestrial ectotherms to climate change, while calling into question if acclimation in marine systems will be able to counteract the potentially negative effects of reductions in relative variance. Assuming changes in relative variation in physiological rates are underpinned by genetic variation, and there is a genetic correlation with fitness, such reductions in physiological variance could limit a populations ability to adapt to future climates [@urban2023and]. Of course, considering how behavioural modifications buffer organisms from high temperatures is also needed. Terrestrial species likely have greater opportunity for behavioural microhabitat selection in thermally heterogeneous environments which is expected to decrease the impacts of climate change [@huey2012predicting; @sunday2014thermal]. Importantly, while we seen a reduction in relative variation, absolute variation in physiological rates did increase with temperature (@fig-s1) -- albeit to a much lesser extent in terrestrial and marine habitats compared to freshwater -- supporting the general notion that capacity for adaptation may still be limited in terrestrial and marine habitats. As of yet, it is unclear whether a focus on relative or absolute variance should be prioritized. Future theoretical models, that vary both mean and variance, are needed to characterise the interplay between changes in mean and variance.

### ***Plasticity and relative variance in physiological rates do not differ between life stages***
We show that early (juvenile) and late (adult) life stages exhibited similar acclimation capacities and changes in relative variance across habitats. The lack of differences we observed may be due to the fact that both juvenile and adult animals occupy similar ecological niches and thermoregulate in similar ways making physiological responses to temperature similar. It is possible that temperatures experienced earlier in development, for example egg or early larval stages, would show different patterns because early developmental constraints are known to affect how physiological systems respond later in life because of programmatic effects [@angilletta2009thermal; @noble2018developmental; @o2019developmental]. Such effects would manifest in juveniles life-history stages

environments are expected Life-history stages often occupy different ecological niches and exhibit different physiological responses, levels of plasticity, and patterns of mortality in response to temperature. As such, it is becoming increasingly important to understand how climate change will impact different life-history stages [@levy2015resolving; @petitgas2013impacts]. Greater plasticity in early life stages of development is expected to be important to increase resilience to the effects of climate change because early life stages are often particularly vulnerable periods in development [@martin2015age; @stearns1976life].  On average, our results show that early life stages are not more vulnerable to the impacts of climate change. However, our general meta-analytic findings do not imply that early life stages are not always less plastic, but that such responses are likely context or trait dependent [@carter2020life; @moghadam2019heat]. For example, @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults. In contrast, @carter2020life showed greater thermal plasticity in metabolic rate for adults but little in pupae of *Onthophagus taurus*.  Despite variation in the literature, very few studies measure multiple life-history stages within the same population. In future research, there needs to be more studies that explicitly compare differences in plasticity among life stages across multiple traits within the same population.
